---
output:
  html_document:
    keep_md: yes
    theme: cerulean
    toc: yes
---
Daily Acitivity Pattern Analysis
========================================================

## Data

- Dataset : [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Research Objectives

- Analyze the daily activity pattern, and compare the result with missing value filled in. 
- Explore if there is difference of the daily activity pattern between weekdays and weekend.

## Data Processing
Change the date and interval variable to time format
```{r readdata}
data <- read.csv("activity.csv", header = TRUE, sep = ",")
data$date_revised <- as.Date(data$date)
data$interval_revised <- sprintf("%04d", data$interval)
data$interval_time <- strptime(data$interval_revised, format = "%H%M")
```


## Analyze daily activity pattern without missing data

### split data by date
```{r totalsteps}
s_date <- split(data, data$date_revised)
total_steps <- sapply(s_date, function(x) sum(x$steps))
```

### Make histogram for total number of steps taken each day
```{r hist_totalstep, fig.height=4}
library(ggplot2)
qplot(total_steps, geom = "histogram", binwidth = 1000, fill = ..count.., xlab = "Total steps per day", ylab = "Number of Days", main = "Total steps per day Frequency")
#plot <- hist(total_steps, main = "Total steps per day", xlab ="total number of steps taken each day")
```

### Calculate mean and median total number of steps taken per day
```{r mean_median_steps}
mean_steps <- mean(total_steps, na.rm = TRUE)
median_steps <- median(total_steps, na.rm = TRUE)
mean_steps
median_steps
```
**mean and median daily steps with out missing value**
mean steps:`r mean_steps`
median steps:`r median_steps`.

###  Plot for average daily activity pattern
```{r seriesplot_activitypattern, fig.height=4,fig.width=8}
library(scales)
s_pattern <- split(data, data$interval_revised) 
average_steps <- sapply(s_pattern, function(x) mean(x$steps, na.rm = TRUE))
day_pattern <- data.frame(time = data$interval_time[1:288], ave_steps = average_steps)
p_day <- ggplot(day_pattern, aes(x = time, y = ave_steps))
p_day+geom_line(colour = "tomato", size = .7)+labs(x = "interval", y = "Number of steps", title = "Average Daily Activity Pattern")+scale_x_datetime(labels= date_format("%H:%M"))
#plot(data$interval_time[1:288], average_steps, type = "l", col = "blue", main = "Average daily activity pattern", xlab = "interval", ylab = "Number of steps")
```

### Get the most active 5-minute interval
```{r mostactive_interval}
max_index <- which(average_steps == max(average_steps))
max_interval <- data$interval_revised[max_index]
max_interval
```
**Most avtive 5-minute interval** `r max_interval` on average across all the days in the dataset, contains the maximum number of steps.

## Explore daily activity pattern with Missing data

### Imputing missing valuse
Calculate and report total number of missing values in the dataset. 
```{r number_missingValue}
bad <- is.na(data$steps)
num_NA <- length(data$steps[bad])
num_NA
```
**Total number of missing values of steps**:  `r num_NA`

Fill all the missing value by the mean for that 5-minute interval

**strategy**

- add a column to dataframe called "steps_revised".
- use mean for 5-minute interval steps as the value in "step_revised" column if it is missing value in original "steps" value. Otherwise, copy the steps number to the "step_revised" column.  
- create a new dataset equal to the original one, but replace the original "step" column's value by the "step_revised" column's value.
```{r fillNA}
data$average_steps <- average_steps
for (i in seq(nrow(data))) {
        if (is.na(data[i, "steps"])) {
                data$steps_revised[i] <- data[i, "average_steps"]
        }
        else {
                data$steps_revised[i] <- data[i, "steps"]
        }
}
```

Create a new dataset with missing data filled in. 
```{r createNewData}
data_revised <- data.frame(steps = data$steps_revised, date = data$date, interval = data$interval)
head(data_revised)
```

### Plot steps taken each day, with missing value filled in
```{r hist_totalsteps_revised, fig.height=4}
s_date_revised <- split(data, data$date_revised)
total_steps_revised <- sapply(s_date_revised, function(x) sum(x$steps_revised, na.rm = TRUE))
qplot(total_steps_revised, geom = "histogram", binwidth = 1000, fill = ..count.., xlab = "Total steps per day", ylab = "Number of Days", main = "Total steps per day(simulated data)")
#plot <- hist(total_steps_revised, main = "Total steps per day_revised", xlab ="total number of steps taken each day_revised")
```

### Calculate mean and median total number of steps taken per day. 
```{r mean_median_revised}
mean_steps <- mean(total_steps_revised)
median_steps <- median(total_steps_revised)
mean_steps
median_steps
```
**mean and median steps,with missiong value filled in**
mean steps:`r mean_steps`
median steps: `r median_steps`.

**Conclusion:** Mean and median value almost the same after we filled in the missing value with the mean steps for that 5-minute interval. Which means missing values don not have big impact on the research result**


## Compare activity patterns between weekdays and weekends
### Create a new factor variable indicating weekday or weekend
```{r factor_dayType}
library(timeDate)
data$day_type <- "weekdays"
data$day_type[isWeekend(data$date_revised)] <- "weekend"
data$day_type <- as.factor(data$day_type)
```

### Calculate average steps of each 5-minutes interval for weekdays and weekend
```{r seriesplot_activitypattern_dayType}
#seperate the weekdays and weekend data into two groups
s_dayType <- split(data, data$day_type)
s_pattern <- lapply(s_dayType, function(x) split(x, x$interval_revised))
data_weekdays <- s_pattern[["weekdays"]]
data_weekend <- s_pattern[["weekend"]]

#calculate the weekdays and weekend's average steps for each 5-minutes interval.
average_steps_weekdays <- sapply(data_weekdays,function(x) mean(x$steps_revised))
average_steps_weekend <- sapply(data_weekend,function(x) mean(x$steps_revised))
```

### Plot activity pattern for weekdays & weekend

```{r, fig.height = 4,fig.width=10}
weekdays <- data.frame(time = data$interval_time[1:288], ave_steps = average_steps_weekdays, day = "weekdays")
weekend <- data.frame(time = data$interval_time[1:288], ave_steps = average_steps_weekend, day = "weekend")
pattern <- rbind(weekdays, weekend)



library(ggplot2)
library(scales)
p <- ggplot(pattern, aes(time, ave_steps,group = day))
p + geom_line(aes(colour = day), size = .7) + labs(x = "Interval", y="Number of steps", title = "Average daily activity pattern")+scale_x_datetime(labels= date_format("%H:%M"))

